Supplemental Materials: \\ \emph{Active contours-driven registration method for the structure-informed segmentation of diffusion MR images}

### 2015-10-24 Publications/2015-NeuroImage/03-introduction.tex
Working on discussion, intro almost finished

\newcomment[RV\#2(C.10)]{%
Here we present a segmentation and surface-to-volume registration method 
  named \regseg{}, and show its usefulness in incorporating the anatomical
  information into native \gls*{dmri} space to aid subsequent reconstruction,
  tractography and network building using a cortical parcellation.
The hypothesis underlying is that the registration and segmentation problems
  in \gls*{dmri} can be solved simultaneously.
To implement \regseg{} we first establish an active-contours without edges
  \citep{chan_active_2001} segmentation framework.
Several \emph{anatomically-correct} surfaces of the same subject initialize
  the 3D active-contours, which evolve searching for homogeneous regions in
  the multivariate target-image.
In the presented application to segment \gls*{dmri}, we propose a bivariate 
  target-volume comprehending the \gls*{fa} and \gls*{adc} maps.
The active-contours are without edges since the \gls*{fa} and \gls*{adc} maps
  do not present steep image gradients (edges) but anatomy can be identified
  by looking for piece-wise smooth homogeneous regions.
The evolution of the surfaces is supported by a B-spline basis, optimized
  iteratively using a descent approach driven by shape-gradients
  \citep{besson_dream2s_2003,herbulot_segmentation_2006}.
Therefore, \regseg{} establishes a registration framework that actually
  overcomes the nonlinear warping induced by \gls*{epi} distortions.%
}
\Regseg{} integrates the benefits of segmentation and registration methods together and
  exploits the multivariate nature of \gls*{dmri} data to contribute in the proposed
  application on three key aspects:
  1) it avoids the resampling the \glspl*{dwi} introduced by the correction of \gls*{epi}
    distortions, since processing can be performed on the native \gls*{dmri} space;
  2) the surfaces are typically extracted from the \gls*{t1} image of the same subject,
    therefore \regseg{} does not require additional MR acquisitions to the minimal
    \gls*{dmri} protocol; and
  3) \regseg{} increases the geometrical accuracy of the overall process.
In this paper, we first verify the functionality of the method and the \regseg{}
  implementation using a set of digital phantoms, demonstrating the sub-pixel accuracy
  of registration.
We then evaluate \regseg{} on real \gls*{dmri} datasets, using a derivation of our
  instrumentation framework \citep{esteban_simulationbased_2014} which simulates 
  known and realistic \gls*{epi} distortions.
We also compare \regseg{} and a nonlinear registration method to map the \lowb{} to 
  the corresponding \gls*{t2} image of the same subject.
This method is a derivation of the \gls*{t2b} correction method, reproducing the settings
  of an existing implementation included within a widely used diffusion processing software
  \citep{leemans_exploredti_2009}.
With this comparison we demonstrate how \regseg{} achieves higher accuracy with the
  simultaneous registration and segmentation process.

>>>>

\newcomment[RV\#2(C.10)]{%
Here we present a segmentation and surface-to-volume registration method 
  named \regseg{}, and show its usefulness in incorporating the anatomical
  information into native \gls*{dmri} space to aid subsequent reconstruction,
  tractography and network building using a cortical parcellation.
The hypothesis underlying is that the registration and segmentation problems
  in \gls*{dmri} can be solved simultaneously.
To implement \regseg{} we first establish an active-contours without edges
  \citep{chan_active_2001} segmentation framework.
Several \emph{anatomically-correct} surfaces of the same subject initialize
  the 3D active-contours, which evolve searching for homogeneous regions in
  the multivariate target-image.
In the presented application to segment \gls*{dmri} data, we propose a bivariate 
  target-volume comprehending the \gls*{fa} and \gls*{adc} maps.
The evolution of the surfaces is supported by a B-spline basis, optimized
  iteratively using a descent approach driven by shape-gradients
  \citep{besson_dream2s_2003,herbulot_segmentation_2006}.
Therefore, \regseg{} establishes a registration framework that actually
  overcomes the nonlinear warping induced by \gls*{epi} distortions.%
}
\Regseg{} integrates the benefits of segmentation and registration methods together and
  exploits the multivariate nature of \gls*{dmri} data to contribute in the proposed
  application on three key aspects:
  1) it avoids the resampling the \glspl*{dwi} introduced by the correction of \gls*{epi}
    distortions, since processing can be performed on the native \gls*{dmri} space;
  2) the surfaces are typically extracted from the \gls*{t1} image of the same subject,
    therefore \regseg{} does not require additional MR acquisitions to the minimal
    \gls*{dmri} protocol; and
  3) \regseg{} increases the geometrical accuracy of the overall process.
In this paper, we first verify the functionality of the method and the \regseg{}
  implementation using a set of digital phantoms, demonstrating the sub-pixel accuracy
  in registration.
We then evaluate \regseg{} on real \gls*{dmri} datasets, using a derivation of our
  instrumentation framework \citep{esteban_simulationbased_2014} which simulates 
  known and realistic \gls*{epi} distortions.
We also compare \regseg{} and a nonlinear registration method to map the \lowb{} to 
  the corresponding \gls*{t2} image of the same subject.
This method is a derivation of the \gls*{t2b} correction method, reproducing the settings
  of an existing implementation included within a widely used diffusion processing software
  \citep{leemans_exploredti_2009}.
With this comparison we demonstrate how \regseg{} achieves higher accuracy with the
  simultaneous registration and segmentation process.

### 2015-10-24 Publications/2015-NeuroImage/06-discussion.tex
Working on discussion, intro almost finished

% -*- root: 00-main.tex -*-
\section{Discussion}
\label{sec:regseg-discussion}

\newcomment[RV\#1(C.3)]{%
\paragraph*{Contributions}\label{sec:regseg-related_work} %
Previously, joint segmentation and registration have been applied successfully to other problems
such as longitudinal object tracking \citep{paragios_level_2003} and atlas-based
  segmentation \citep{gorthi_active_2011}.
The most common approach to solve the problem simultaneously is optimizing a deformation model
  (registration) that supports the evolution the active contours (segmentation), like
  \citep{paragios_level_2003,yezzi_variational_2003}.
\cite{unal_coupled_2005}, and later \cite{wang_joint_2006}, replaced the linear registration
  transform of \cite{yezzi_variational_2003} with a free-form deformation field.
\cite{droske_mumfordshah_2009} reviewed the existing techniques and proposed two different
  approaches for applying the Mumford-Shah functional \citep{mumford_optimal_1989} during simultaneous
  registration and segmentation by propagating the deformation field from
  the contours onto the whole image definition.
Finally, \cite{gorthi_active_2011} extended the existing methodologies using a multiphase
  level set function to register several active contours during the application
  of atlas-based segmentation.
Unlike the aforementioned registration-and-segmentation methods, \regseg{} uses shape-gradients
  \citep{herbulot_segmentation_2006} computed on a hierarchical set of explicit surfaces
  (triangularized meshes) that substitute the multiphase level sets.
Recently, \cite{guyader_combined_2011} proposed a simultaneous segmentation and
  registration method in 2D using level sets and a nonlinear elasticity smoother on the
  displacement vector field, which preserves the topology even with very large deformations.
\Regseg{} includes an anisotropic regularizer for the displacements field proposed by
  \cite{nagel_investigation_1986}.
This regularization strategy conceptually falls in the midway between the Gaussian smoothing
  generally included in most of the existing methodologies, and the complexity of
  the elasticity smoother of \cite{guyader_combined_2011}.

An important antecedent of \regseg{} is \emph{bbregister} \citep{greve_accurate_2009}.
The tool has been widely adopted as the standard registration method to be used along with the \gls*{epi}
  correction of choice.
It implements a linear mapping, and uses 3D active contours with edges\footnote{A three-dimensional active contour
  is a surface and it is ``with edges'' when the contours evolve in search of abrupt intensity steps
  (the edges) within the target image.} to search for intensity boundaries in the \lowb{} image.
The active contours are initialized using surfaces extracted from the \gls*{t1} using
  \emph{FreeSurfer} \citep{fischl_freesurfer_2012}: the \emph{pial} surface (exterior of the
  cortical \gls*{gm}) and the \emph{white} surface (the \gls*{wm}/\gls*{gm} interface).
The \lowb{} image only includes a detectable frontier for the pial surface, and thus
  \emph{bbregister} is limited to aligning the cortical layer in this
  application.
To overcome the problem of nonlinear distortions, \emph{bbregister} excludes the
  regions that are typically warped from the boundary search.
Indeed, the distortion must be addressed separately because it is not supported by
  the affine transformation model.
\Regseg{} can be seen as a derivative of \emph{bbregister} for the use of active surfaces.
However, the deformation model is nonlinear and the surfaces in \regseg{} look for the
  homogeneity within the regions enclosed, instead of image edges.}

@@

\paragraph*{Accuracy tests}
The hypothesis tested in our study was that reliable image registration can be performed
  by searching for the homogeneous regions defined by a set of nested surfaces in a
  multispectral image
The surfaces correspond to precise contours extracted from another image of the same subject
  (i.e. a \gls*{t1} image, or a different time step).
We demonstrated that active contours without edges can be used successfully to drive a
  deformation supported by B-spline basis functions with digital phantoms.
We randomly deformed four different phantom models to mimic three homogeneous regions
  (\gls*{wm}, \gls*{gm}, and \acrlong*{csf}) and we used them to simulate \gls*{t1} and \gls*{t2}
  images at two resolution levels.
After registration with \regseg{}, we measured the Hausdorff distance between the
  projected contours obtained using the ground-truth warping and our estimates.
We concluded that the errors were significantly lower than the voxel resolution.
We also assessed the 95\% \gls*{ci}, which yielded an aggregate interval of
  0.64--0.66 [mm] for the low resolution phantoms (2.0 mm isotropic voxel) and
  0.34--0.38 [mm] for the high resolution phantoms (1.0 mm isotropic).
Therefore, we also concluded that the error was bounded above by half of the
  voxel spacing.
\newcomment[RV\#1(C.8)]{%}
The distributions of errors along surfaces vary importantly depending on the shape of the
  phantom (see \autoref{fig:regseg-phantom}B).
The misregistration error of the ``gyrus'' phantom showed a much lower spread than that
  for the other shapes.
We argue that the symmetry of those other shapes posed difficulties in driving the contours
  towards the appropriate region and producing some ``sliding'' effect between the
  faces of the surfaces and their ground-truth position.
This effect should not be present in real datasets, thanks to the very convoluted cortical
  layer, and the directional restriction of distortion.}

\paragraph*{Application to real data}
We designed \regseg{} as a method for segmenting \gls*{dmri} data by exploiting the
  anatomical information extracted from a structural image (typically a \gls*{t1} image)
  of the subject.
Current pipelines using whole-brain tractography usually solve this problem with a two-step approach.
First, the images are corrected for nonlinear distortions using auxiliary acquisitions
  such as fieldmaps \citep{jezzard_correction_1995}, \glspl*{dwi} with reversed \gls*{pe}
  blips \citep{chiou_simple_2000}, or \gls*{t2} images \citep{kybic_unwarping_2000}.
Second, the segmentation is projected from a reference \gls*{t1} image using linear
  registration \citep{greve_accurate_2009}.
\Regseg{} addresses this joint problem in a single step and it does not require any additional
  acquisition other than the minimal protocol using only \gls*{t1} and \gls*{dmri} images.
This situation is found commonly in historical datasets.
Moreover, since the structural information is projected into the native space of \gls*{dmri}
  reconstruction and tractography can be performed without resampling data to an undistorted
  space.

>>>>

% -*- root: 00-main.tex -*-
\section{Discussion}
\label{sec:regseg-discussion}
We present \regseg{}, a simultaneous segmentation and registration method that
  maps a set of nested surfaces into a multivariate target-image.
The nonlinear registration process evolves driven by the fitness of the
  picewise-smooth classification of voxels in the target volume imposed
  by the current mapping of the surfaces.
We propose \regseg{} to incorporate anatomical information extracted from \gls*{t1}
  images into the corresponding \gls*{dmri} of the same subject.
Previously, joint segmentation and registration have been applied successfully to
  other problems such as longitudinal object tracking \citep{paragios_level_2003}
  and atlas-based segmentation \citep{gorthi_active_2011}.
The most common approach to solve the problem simultaneously is optimizing a deformation
  model (registration) that supports the evolution the active contours (segmentation),
  like \cite{paragios_level_2003,yezzi_variational_2003}.
\Regseg{} can be seen as particular case of atlas-based segmentation, replacing the atlas
  by the structural image of the subject.
However, the majority of atlas-based segmentation methods, such as \citep{gorthi_active_2011},
  use a multiphase level set function to register several active contours.
Alternatively, \regseg{} implements the active contours with a hierarchical set of explicit
  surfaces (triangular meshes) that substitute the multiphase level sets, and registration
  is driven by shape-gradients \citep{herbulot_segmentation_2006}.

An important antecedent of \regseg{} is \emph{bbregister} \citep{greve_accurate_2009}.
The tool has been widely adopted as the standard registration method to be used along with the \gls*{epi}
  correction of choice.
It implements a linear mapping, and uses 3D active contours \emph{with edges} to
  search for intensity boundaries in the \lowb{} image.
The active contours are initialized using surfaces extracted from the \gls*{t1} using
  \emph{FreeSurfer} \citep{fischl_freesurfer_2012}: the \emph{pial} surface (exterior of the
  cortical \gls*{gm}) and the \emph{white} surface (the \gls*{wm}/\gls*{gm} interface).
The \lowb{} image only includes a detectable frontier for the pial surface, and thus
  \emph{bbregister} is limited to aligning the cortical layer in this
  application.
To overcome the problem of nonlinear distortions, \emph{bbregister} excludes the
  regions that are typically warped from the boundary search.
Indeed, the distortion must be addressed separately because it is not supported by
  the affine transformation model.
The deformation model of \regseg{} is nonlinear and the surfaces are active
  contours \emph{without edges} \citep{chan_active_2001}.
The contours are without edges because the \gls*{fa} and \gls*{adc} maps do
  not present steep image gradients (edges) but the anatomy can be identified
  by looking for piece-wise smooth homogeneous regions.

Recently, \cite{guyader_combined_2011} proposed a simultaneous segmentation and
  registration method in 2D using level sets and a nonlinear elasticity smoother on the
  displacement vector field, which preserves the topology even with very large deformations.
\Regseg{} includes an anisotropic regularizer for the displacements field proposed by
  \cite{nagel_investigation_1986}.
This regularization strategy conceptually falls in the midway between the Gaussian smoothing
  generally included in most of the existing methodologies, and the complexity of
  the elasticity smoother of \cite{guyader_combined_2011}.
Other minor features that differ from current methods in joint segmentation and registration are
  the support of multivariate target-images and the efficient computation of the shape-gradients
  implemented with sparse matrices.

We verified that precise segmentation and registration of a set of surfaces into multivariate
  data is possible on digital phantoms.
We randomly deformed four different phantom models to mimic three homogeneous regions
  (\gls*{wm}, \gls*{gm}, and \acrlong*{csf}) and we used them to simulate \gls*{t1} 
  and \gls*{t2} images at two resolution levels.
After registration with \regseg{}, we measured the Hausdorff distance between the
  projected contours obtained using the ground-truth warping and our estimates.
We concluded that the errors were significantly lower than the voxel resolution.
We also assessed the 95\% \gls*{ci}, which yielded an aggregate interval of
  0.64--0.66 [mm] for the low resolution phantoms (2.0 mm isotropic voxel) and
  0.34--0.38 [mm] for the high resolution phantoms (1.0 mm isotropic).
Therefore, we also concluded that the error was bounded above by half of the
  voxel spacing.
\newcomment[RV\#1(C.8)]{%}
The distributions of errors along surfaces vary importantly depending on the shape of the
  phantom (see \autoref{fig:regseg-phantom}B).
The misregistration error of the ``gyrus'' phantom showed a much lower spread than that
  for the other shapes.
We argue that the symmetry of those other shapes posed difficulties in driving the contours
  towards the appropriate region and producing some ``sliding'' effect between the
  faces of the surfaces and their ground-truth position.
This effect should not be present in real datasets, thanks to the very convoluted cortical
  layer, and the directional restriction of distortion.}

\todo{real data}
This application is necessary in the extraction of structural connectivity from
  \gls*{dmri}, and has been typically solved in a two-step approach.
First, the \glspl*{dwi} are corrected for \emph{\gls*{epi} distortions} by estimating
  the nonlinear-deformation field from extra MR acquisitions
  \citep{jezzard_correction_1995,chiou_simple_2000,cordes_geometric_2000,
  kybic_unwarping_2000}.
Second, mapping the structural information from the corresponding \gls*{t1} image
  using a linear registration tool like \emph{bbregister} \citep{greve_accurate_2009}.
numerous attempts to delineate the structures of interest in diffusion
  space have been proposed.
Current activity on improving correction methods \citep{irfanoglu_drbuddi_2015} and
  segmenting the \gls*{dmri} in its native space \citep{jeurissen_tissuetype_2015}
  proofs the open interest of the application.
\todo{real data old}
We designed \regseg{} as a method for segmenting \gls*{dmri} data by exploiting the
  anatomical information extracted from a structural image (typically a \gls*{t1} image)
  of the subject.
Current pipelines using whole-brain tractography usually solve this problem with a two-step approach.
First, the images are corrected for nonlinear distortions using auxiliary acquisitions
  such as fieldmaps \citep{jezzard_correction_1995}, \glspl*{dwi} with reversed \gls*{pe}
  blips \citep{chiou_simple_2000}, or \gls*{t2} images \citep{kybic_unwarping_2000}.
Second, the segmentation is projected from a reference \gls*{t1} image using linear
  registration \citep{greve_accurate_2009}.
\Regseg{} addresses this joint problem in a single step and it does not require any additional
  acquisition other than the minimal protocol using only \gls*{t1} and \gls*{dmri} images.
This situation is found commonly in historical datasets.
Moreover, since the structural information is projected into the native space of \gls*{dmri}
  reconstruction and tractography can be performed without resampling data to an undistorted
  space.

\newcomment[RV\#1(C.3)]{%
\paragraph*{Contributions}\label{sec:regseg-related_work} %

}

@@
